Load data (hidden)

```{r, echo=FALSE,results=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
source("common.R")

## guess Ag labels from covid from earlier version
## (load(file_rdata_v1))
## v1 <- m
## with(v1,table(group,type))
## covid.v1 <- m[type=="COVID"]

## load v3 data for comparison
library(readxl)
(nm <- excel_sheets(file_xlsx_v3))
data <- lapply(seq_along(nm), function(i) {
  x <- read_xlsx(file_xlsx_v3,sheet=i)  %>% as.data.table()
  setnames(x,make.names(names(x)))
  x$OD  %<>% as.numeric()
  x$nm <- nm[i]
  x
  ## message(i)
  ## print(sapply(x,class))
})
names(data) <- make.names(nm)
lapply(data, names) ##check
setnames(data[[6]],"Antigen","Ag")
## some pairs are in same order - exploit that
for(i in c(2,3,4,5,6)) {
  data[[i]][,nm:=nm[i]][,rowid:=1:.N]
}
x3=data[-1]

## load v4 data - just take newest data
library(readxl)
(nm <- excel_sheets(file_xlsx_v4))
## data <- lapply(seq_along(nm), function(i) {
## check all data with assay date before july already represented
x <- read_xlsx(file_xlsx_v4,sheet=1)  %>% as.data.table()
setnames(x,make.names(names(x)))
old.ids <- lapply(data, "[[", "Sample.ID")  %>% unlist()  %>% unique()
table(x[Assay.date!="200702"]$Sample.ID %in% old.ids) # all TRUE

## extract 200702 data
x <- read_xlsx(file_xlsx_v4,sheet=2)  %>% as.data.table()
setnames(x,make.names(names(x)))
table(x$Assay.date)
x$OD  %<>% as.numeric()
x$nm <- "200702"
x[Sampling.date==206011,Sampling.date:=200611] # typo
x[,redo:=Plate %in% c(1,2,3,7,8,9) & Ag=="SPIKE" ]

## add rerun for S, plates 1,2,3,7,8,9
(nm <- excel_sheets(file_xlsx_v5))
## data <- lapply(seq_along(nm), function(i) {
## check all data with assay date before july already represented
x5 <- read_xlsx(file_xlsx_v5,sheet=1)  %>% as.data.table()
setnames(x5,make.names(names(x5)))
head(x5)

head(x[Plate==1][redo==TRUE])
head(x5[Plate==1])

## layout has changed
x5[Plate==3,Sample.ID:=sub("WK24","Wk24",Sample.ID)]

for(pl in c(1:3,7:9)) {
  message("\nPlate ",pl)
message("on new, not old")
print(setdiff(x5[Plate==pl]$Sample.ID,x[Plate==pl]$Sample.ID))
message("\non old, not new")
print(setdiff(x[Plate==pl]$Sample.ID,x5[Plate==pl]$Sample.ID))
message("\nnumber of No S C, old/new: ",
        nrow(x[Plate==pl & Sample.ID=="No S C"]), "/",
        nrow(x5[Plate==pl & Sample.ID=="No S C"]))
}

x5$OD  %<>% as.numeric()
x5$nm <- "200702"

## add weeks 30-34 data - x6 superceded by v7
## add rerun for plates YP, YO and SP replaced with plates ÅÅ, ÅP, ÅQ, and inc. some more negative controls.
x7 <- read_xlsx(file_xlsx_v7,sheet=1)  %>% as.data.table()
setnames(x7,make.names(names(x7)))
head(x7)
x7$OD  %<>% as.numeric()
x7$nm <- "200923"

## v8 - 
## I ran an additional 33 individuals who are PCR+ staff members at the hospital, and some controls. 
## I have run another 300 spring 2019 controls for S and RBD, attached. 

x8a=read_xlsx(file_xlsx_v8[[1]],sheet=1) %>% as.data.table()
x8b=read_xlsx(file_xlsx_v8[[2]],sheet=1) %>% as.data.table()
names(x8a)[1]="Sample ID" # funny chararcter in "Sample ID "
x8=rbind(x8a,x8b,fill=TRUE)
setnames(x8, make.names(names(x8)))
setnames(x8, "Antigen","Ag") # to match previous
setnames(x8, "Collection.date","Sampling.date") # to match previous
x8$OD  %<>% as.numeric()
x8[,nm:="201007"]
x8[is.na(Assay.date),Assay.date:=201007] # some entries missing

## x9 1612
(nm <- excel_sheets(file_xlsx_v9))
## data <- lapply(seq_along(nm), function(i) {
## check all data with assay date before july already represented
x9 <- read_xlsx(file_xlsx_v9,sheet=1)  %>% as.data.table()
setnames(x9,make.names(names(x9)))
setnames(x9,"Sample.ID.","Sample.ID")
setnames(x9,"Antigen","Ag")
setnames(x9,"Collection.date","Sampling.date")
x9[,nm:="201216"]
x9[,OD:=as.numeric(OD)]
head(x9)

## combine
xcomb <- rbind(x[redo==FALSE][,redo:=NULL], x5, x7,x8, x9,fill=TRUE)
data <- c(x3, list(xcomb))  %>% rbindlist(., fill=TRUE)

## label nicely
m <- copy(data)
sample(m$Sample.ID,100)
m[,group:=sub(" .*","",nm)]
m[Sample.ID %in% c("No S C","No S Ctrl","NO S C"), type:="No sample control"]
m[,Sample.ID:=sub("wk","Wk",Sample.ID)][,Sample.ID:=sub("WK","Wk",Sample.ID)]
m[Sample.ID=="P4", type:="Patient 4"]
m[Status %in% c("COVID PCR+","COVID+"),type:="COVID"]
m[grepl("^P Wk|^p Wk|^P WK",Sample.ID), type:="Pregnant volunteers"] # fix case typo
m[grepl("^Wk",Sample.ID), type:="Blood donors"]
m[grepl("^Sp2019|B1q",Sample.ID), type:="Historical controls"]
m[type %in% c("Pregnant volunteers","Historical donors"),Sample.ID:=gsub("WK|wk","Wk",Sample.ID)]
m[,stype:=factor(type)][,stype:=relevel(stype,"Historical controls")]
## with(m[is.na(type)], sample(Sample.ID,100))
m[is.na(type) & grepl("Spring2019",Sample.ID), type:="Historical controls"]
m[is.na(type) & grepl("Spring2019",Sample.ID), type:="Historical controls"]
with(m[is.na(type)], table(Sampling.date))
head(m[is.na(type)])
tail(m[is.na(type)])
m[!is.na(Assay.date),
  adate:=as.Date(paste0("20",Assay.date),format="%Y%m%d")]
head(m[Plate==1])
tail(m[Plate==1])
m[grep("Wk",Sample.ID), Week:=sub(".*(Wk[0-9]+).*","\\1",Sample.ID) ]
table(m$Week)
```


## adjust OD by no sample controls

examine outlier in spike control
```{r}

m[type=="No sample control" & Plate=="!",.(Ag,type,Plate,OD)][order(Ag,OD)]
head(sort(m[type=="No sample control"]$OD))
```

calculate mean of non-outlier controls and normalise, then reshape wide
```{r}
norm.plate <- m[type=="No sample control" & OD>0.03, # one outlier control
                .(control.plate=mean(OD)),by=c("Ag","Plate")]
norm.date <- m[type=="No sample control" & OD>0.03, # one outlier control
               .(control.date=mean(OD)),by=c("Ag","adate")]
m <- merge(m,norm.plate,by=c("Ag","Plate"),all.x=TRUE)
m <- merge(m,norm.date,by=c("Ag","adate"),all.x=TRUE)

m[,NOD:=OD*mean(norm.plate$control.plate) /
     ifelse(is.na(control.plate),control.date,control.plate)]

## no more need of no s c
m <- m[type!="No sample control"]
mm1 <- merge(m[group %in% c("BD","Preg") & Ag=="SPIKE" & type!="No sample control"],
            m[group %in% c("BD","Preg") & Ag=="RBD" & type!="No sample control"],
            by=c("group","Sample.ID","rowid","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid:=NULL]
mm2 <- merge(m[group %in% c("COVID19") & Ag=="SPIKE" & type!="No sample control"],
            m[group %in% c("COVID19") & Ag=="RBD" & type!="No sample control"],
            by=c("group","Sample.ID","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid.spike:=NULL][,rowid.rbd:=NULL]
mm3 <- merge(m[group %in% c("200702") & Ag=="SPIKE" & type!="No sample control"],
            m[group %in% c("200702") & Ag=="RBD" & type!="No sample control"],
            by=c("group","Sample.ID","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid.spike:=NULL][,rowid.rbd:=NULL]
mm4 <- merge(m[group %in% c("200923") & Ag=="SPIKE" & type!="No sample control"],
            m[group %in% c("200923") & Ag=="RBD" & type!="No sample control"],
            by=c("group","Sample.ID","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid.spike:=NULL][,rowid.rbd:=NULL]
mm5 <- merge(m[group %in% c("201007") & Ag=="SPIKE" & type!="No sample control"],
            m[group %in% c("201007") & Ag=="RBD" & type!="No sample control"],
            by=c("group","Sample.ID","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid.spike:=NULL][,rowid.rbd:=NULL]
mm6 <- merge(m[group %in% c("201216") & Ag=="SPIKE"],
            m[group %in% c("201216") & Ag=="RBD"],
            by=c("group","Sample.ID","type","stype"),
            suffixes=c(".spike",".rbd"))[,rowid.spike:=NULL][,rowid.rbd:=NULL]
mm <- rbind(mm1,mm2,mm3,mm4,mm5,mm6)
setnames(mm,c("NOD.rbd","NOD.spike"),c("RBD","SPIKE"))

plot.adjusted <- ggplot(mm[type!="No sample control"][order(stype,decreasing=TRUE)],
                        aes(x=SPIKE,y=RBD,col=type)) +
  geom_point(alpha=0.7) + # strongly correlated
  ## geom_density2d() +
  facet_wrap(~group) +
  background_grid() +
  scale_x_log10() +
  scale_y_log10()+ theme(legend.position="bottom") + ggtitle("Adjusted OD")
plot.adjusted
m <- mm
```


## count
Table of sample numbers in each class
```{r}
with(m,table(group,type,exclude=NULL))
with(m,table(assay.date=adate.rbd,type,exclude=NULL))
with(m,table(assay.date=adate.spike,type,exclude=NULL))
```

## check distributions match those from Xaq
```{r}
plot_grid(ggplot(m, aes(x=group,col=type,y=OD.spike)) + geom_boxplot() + theme(legend.position="bottom"),
          ggplot(m, aes(x=group,col=type,y=OD.rbd)) + geom_boxplot() + theme(legend.position="bottom"))
```

SPIKE and RBD are correlated in raw data too

```{r,echo=FALSE}
plot.raw <- ggplot(m[type!="No sample control"][order(stype,decreasing=TRUE)],
                   aes(x=OD.spike,y=OD.rbd,col=type)) +
  geom_point(alpha=0.7) + # strongly correlated
  ## geom_density2d() +
  facet_wrap(~group) +
  ## facet_grid(group~type) +
  background_grid() +
  scale_x_log10() +
  scale_y_log10()+ theme(legend.position="bottom") + ggtitle("Raw OD")
plot.raw

```
normalisation improves coefficient of variation (COV) for patient samples, but not historical controls - because these are just noise?
```{r}
m[type=="Patient 4",.(Ag="SPIKE",
                      raw=sd(log(OD.spike))/mean(log(OD.spike)),
                      adj=sd(log(SPIKE))/mean(log(SPIKE))),by="type"]
m[type=="Patient 4",.(Ag="RBD",
                      raw=sd(log(OD.rbd))/mean(log(OD.rbd)),
                      adj=sd(log(RBD))/mean(log(RBD))),by="type"]
m[type=="COVID",.(Ag="SPIKE",
                      raw=sd(log(OD.spike))/abs(mean(log(OD.spike))),
                      adj=sd(log(SPIKE))/abs(mean(log(SPIKE)))),by="type"]
m[type=="COVID",.(Ag="RBD",
                      raw=sd(log(OD.rbd))/abs(mean(log(OD.rbd))),
                      adj=sd(log(RBD))/abs(mean(log(RBD)))),by="type"]
m[type=="Historical controls",.(Ag="SPIKE",
                      raw=sd(log(OD.spike))/abs(mean(log(OD.spike))),
                      adj=sd(log(SPIKE))/abs(mean(log(SPIKE)))),by=c("type","group")]
m[type=="Historical controls",.(Ag="RBD",
                      raw=sd(log(OD.rbd))/abs(mean(log(OD.rbd))),
                      adj=sd(log(RBD))/abs(mean(log(RBD)))),by=c("type","group")]
```



However, adjustment improves comparison of distributions for same
samples in blood/pregnant donor groups

```{r}
with(m[type=="Historical controls"], kruskal.test(OD.spike,group))
with(m[type=="Historical controls"], kruskal.test(SPIKE,group))
with(m[type=="Historical controls"], kruskal.test(OD.rbd,group))
with(m[type=="Historical controls"], kruskal.test(RBD,group))
```

but doesn't improve individual-level correlation in controls - again presumably because these are true zeros, and noise shouldn't correlate between experiments

```{r}
fwrite(m[type=="COVID",.(Sample.ID,adate.spike,Plate.spike,adate.rbd,Plate.rbd)],
       file="~/covid.csv")

library(GGally)
tmp <- merge(m[type=="Historical controls" & group=="BD"],
             m[type=="Historical controls" & group=="Preg"],
             by="Sample.ID")
p1 <- ggpairs(tmp[,.(OD.spike.x,OD.spike.y,OD.rbd.x,OD.rbd.y)]) + ggtitle("Raw")
p3 <- ggpairs(tmp[,.(SPIKE.x,SPIKE.y,RBD.x,RBD.y)]) + ggtitle("Normalised")
plot_grid(
  ggmatrix_gtable(p1),
  ggmatrix_gtable(p3),
  nrow = 1
)
cor(log(tmp$OD.spike.x), log(tmp$OD.spike.y))
cor(log(tmp$SPIKE.x), log(tmp$SPIKE.y))
cor(log(tmp$OD.rbd.x), log(tmp$OD.rbd.y))
cor(log(tmp$RBD.x), log(tmp$RBD.y))

## correlation within plates
tmp[,.(N=.N,cr=cor(SPIKE.x,SPIKE.y)),by=c("Plate.spike.x","Plate.spike.y")]
tmp[,.(cor(RBD.x,RBD.y)),by=c("Plate.rbd.x","Plate.rbd.y")]

```

## add folds, ensuring that samples from the same individual fall into the same fold
```{r}
set.seed(42)
muniq <- m[!duplicated(Sample.ID) & type %in% c("Historical controls","COVID"),
           .(Sample.ID, type, group, rowid)]

## stratified folds
muniq <- muniq[sample(1:.N)]
muniq[,fold1:=1:.N %% 10 + 1,by="type"]
muniq <- muniq[sample(1:.N)]
muniq[,fold2:=1:.N %% 10 + 1,by="type"]
muniq <- muniq[sample(1:.N)]
muniq[,fold3:=1:.N %% 10 + 1,by="type"]

## random folds
muniq <- muniq[sample(1:.N)]
muniq[,fold4:=1:.N %% 10 + 1]
muniq <- muniq[sample(1:.N)]
muniq[,fold5:=1:.N %% 10 + 1]
muniq <- muniq[sample(1:.N)]
muniq[,fold6:=1:.N %% 10 + 1]

## unbalanced folds
muniq[type=="Historical controls",fold7:=sample(1:10,.N,replace=TRUE)]
muniq[type=="COVID",fold7:=sample(1:10,.N,prob=sqrt(1:10)/55,replace=TRUE)]
muniq[type=="Historical controls",fold8:=sample(1:10,.N,replace=TRUE)]
muniq[type=="COVID",fold8:=sample(1:10,.N,prob=sqrt(1:10)/55,replace=TRUE)]
muniq[type=="Historical controls",fold9:=sample(1:10,.N,replace=TRUE)]
muniq[type=="COVID",fold9:=sample(1:10,.N,prob=sqrt(1:10)/55,replace=TRUE)]

table(muniq$fold1,muniq$type)
table(muniq$fold4,muniq$type)
table(muniq$fold7,muniq$type)

m  %<>% cbind(., muniq[match(m$Sample.ID, Sample.ID), paste0("fold",1:9), with=FALSE])

## rename for compatability
m[group=="BD",group:="Blood Donors"]
m[group=="Preg",group:="Pregnant volunteers"]
m[group=="COVID19",group:="COVID+"]

m <- m[,.(group,type,Sample.ID,OD.spike,OD.rbd,SPIKE,RBD,fold1,fold2,fold3,fold4,fold5,fold6,fold7,fold8,fold9)]
## save
save(m, file=file_rdata, version=2)
save(m, file="adjusted-data.RData", version=2)
```
